---
title: "Class 12: Genome Informatics"
author: "Raquel Gonzalez (A16207442)"
format: html
editor: visual
---

# Section 1. Proportion og G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39830003-39921005;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) * 100
```

## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether thre is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3 expression.

How many samples do we have?

```{r}
expr <- read.table("expression_genotype_results.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
medians <- aggregate(exp ~ geno, data = expr, FUN = median)
medians
```

The sample size for A/A is 108 individuals, the sample size for A/G is 233 individuals, and the sample size for G/G is 121 individals. The median expression levels for A/A is 31.25, for A/G is 25.06, and for G/G is 20.07. 

Let's make a boxplot to communicate our results. 

```{r}
library(ggplot2)
ggplot(expr) +
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE) 
```

The relative expression level from A/A suggests an increased expression of this gene, whereas the one from G/G suggests a decreased expression of this gene. 
The SNP does impact the expression of ORMDL3.